Imputation of Incomplete Multilevel Data with \pkg{mice}

### 2022-10-07 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex
Update to work on HIV example together

@@

@@

\begin{CodeChunk}
\begin{CodeInput}
R> set.seed(2022)
R> library(mice)         # for imputation
R> library(ggmice)       # for visualization
R> library(ggplot2)      # for visualization
R> library(dplyr)        # for data wrangling
R> library(lme4)         # for multilevel modeling
R> library(mitml)        # for multilevel pooling
\end{CodeInput}
\end{CodeChunk}

@@

             Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
(Intercept)     4.820     0.353    13.648     4.245     0.000    33.200     0.979 
sex             0.971     0.376     2.581     4.128     0.059    63.141     0.989

                            Estimate 
Intercept~~Intercept|school    0.259 
Residual~~Residual             1.065 
ICC|school                     0.197

@@

             Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
(Intercept)     5.121     0.390    13.125     4.242     0.000    33.503     0.979 
sex             0.673     0.418     1.611     4.121     0.180    66.343     0.989

                            Estimate 
Intercept~~Intercept|school    0.317 
Residual~~Residual             1.253 
ICC|school                     0.213

@@

             Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
(Intercept)     4.941     0.422    11.708     4.203     0.000    39.844     0.982 
sex             1.029     0.537     1.916     4.076     0.127   105.671     0.993

                            Estimate 
Intercept~~Intercept|school    0.306 
Residual~~Residual             1.308 
ICC|school                     0.195

@@

\hypertarget{case-study-ii-hiv-data}{%
\section{Case study II: HIV data}\label{case-study-ii-hiv-data}}

{[}TODO: change the order of case study 2 and 3{]}

\hypertarget{case-study-iii-impact-data}{%
\section{Case study III: IMPACT data}\label{case-study-iii-impact-data}}

@@

\begin{itemize}
\tightlist
\item
  \texttt{name} Name of the study,
\item
  \texttt{type} Type of study (RCT: randomized controlled trial, OBS:
  observational cohort),
\item
  \texttt{age} Age of the patient,
\item
  \texttt{motor\_score} Glasgow Coma Scale motor score,
\item
  \texttt{pupil} Pupillary reactivity,
\item
  \texttt{ct} Marshall Computerized Tomography classification,
\item
  \texttt{hypox} Hypoxia (0=no, 1=yes),
\item
  \texttt{hypots} Hypotension (0=no, 1=yes),
\item
  \texttt{tsah} Traumatic subarachnoid hemorrhage (0=no, 1=yes),
\item
  \texttt{edh} Epidural hematoma (0=no, 1=yes),
\item
  \texttt{mort} 6-month mortality (0=alive, 1=dead).
\end{itemize}

@@

\hypertarget{complete-case-analysis}{%
\subsection{Complete case analysis}\label{complete-case-analysis}}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> fit <- glmer(mod, family = "binomial", data = na.omit(dat)) # fit the model
R> tidy(fit, conf.int = TRUE, exponentiate = TRUE)             # print estimates
\end{CodeInput}
\begin{CodeOutput}
# A tibble: 11 x 9
   effect  group term  estimate std.error statistic   p.value conf.low conf.high
   <chr>   <chr> <chr>    <dbl>     <dbl>     <dbl>     <dbl>    <dbl>     <dbl>
 1 fixed   <NA>  (Int~   0.0863   0.0182     -11.6   2.99e-31   0.0571     0.130
 2 fixed   <NA>  type~   0.757    0.137       -1.54  1.22e- 1   0.531      1.08 
 3 fixed   <NA>  age     1.03     0.00265     12.9   7.40e-38   1.03       1.04 
 4 fixed   <NA>  moto~   0.651    0.0732      -3.82  1.34e- 4   0.522      0.811
 5 fixed   <NA>  moto~   0.489    0.0555      -6.30  2.97e-10   0.391      0.611
 6 fixed   <NA>  moto~   0.274    0.0321     -11.0   2.28e-28   0.218      0.345
 7 fixed   <NA>  pupi~   3.20     0.317       11.7   8.18e-32   2.63       3.88 
 8 fixed   <NA>  pupi~   1.75     0.195        5.06  4.27e- 7   1.41       2.18 
 9 fixed   <NA>  ctIII   2.41     0.268        7.89  3.05e-15   1.94       2.99 
10 fixed   <NA>  ctIV~   2.30     0.214        8.95  3.55e-19   1.92       2.76 
11 ran_pa~ name  sd__~   0.230   NA           NA    NA         NA         NA    
\end{CodeOutput}
\end{CodeChunk}

@@

Create a methods vector and predictor matrix, and make sure
\texttt{name} is not included as predictor:

@@

>>>>

TODO: send complete manuscript by september 30th!

@@

TODO: explicit statement about not going into workings of the methods.
Galimer 2l methods.

@@

\begin{CodeChunk}
\begin{CodeInput}
R> set.seed(2022)
R> library(mice)         # for imputation
R> library(miceadds)     # for imputation
R> library(ggmice)       # for visualization
R> library(ggplot2)      # for visualization
R> library(dplyr)        # for data wrangling
R> library(lme4)         # for multilevel modeling
R> library(mitml)        # for multilevel pooling
\end{CodeInput}
\end{CodeChunk}

@@

             Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
(Intercept)     5.010     0.316    15.841     4.324     0.000    25.159     0.972 
sex             0.748     0.413     1.812     4.112     0.142    72.195     0.990

                            Estimate 
Intercept~~Intercept|school    0.271 
Residual~~Residual             1.125 
ICC|school                     0.213

@@

             Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
(Intercept)     5.043     0.293    17.187     4.469     0.000    17.558     0.961 
sex             0.799     0.329     2.428     4.206     0.069    39.298     0.982

                            Estimate 
Intercept~~Intercept|school    0.335 
Residual~~Residual             1.296 
ICC|school                     0.208

@@

             Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
(Intercept)     5.051     0.259    19.541     4.645     0.000    12.876     0.947 
sex             0.820     0.235     3.484     4.426     0.021    19.256     0.964

                            Estimate 
Intercept~~Intercept|school    0.350 
Residual~~Residual             1.318 
ICC|school                     0.217

@@

\hypertarget{case-study-ii-impact-data-syst-missingness-pred-matrix}{%
\section{Case study II: IMPACT data (syst missingness, pred
matrix)}\label{case-study-ii-impact-data-syst-missingness-pred-matrix}}

@@

\begin{itemize}
\tightlist
\item
  \texttt{name} Name of the study,
\item
  \texttt{type} Type of study (RCT: randomized controlled trial, OBS:
  observational cohort),
\item
  \texttt{age} Age of the patient,
\item
  \texttt{motor\_score} Glasgow Coma Scale motor score,
\item
  \texttt{pupil} Pupillary reactivity,
\item
  \texttt{ct} Marshall Computerized Tomography classification, {[}TODO:
  make this one var? also shows that you don't always need random
  effects everywhere?{]}
\item
  \texttt{hypox} Hypoxia (0=no, 1=yes),
\item
  \texttt{hypots} Hypotension (0=no, 1=yes),
\item
  \texttt{tsah} Traumatic subarachnoid hemorrhage (0=no, 1=yes),
\item
  \texttt{edh} Epidural hematoma (0=no, 1=yes),
\item
  \texttt{mort} 6-month mortality (0=alive, 1=dead).
\end{itemize}

@@

\hypertarget{complete-case-analysis-todo-remove-this}{%
\subsection{Complete case analysis {[}TODO: remove
this?{]}}\label{complete-case-analysis-todo-remove-this}}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> fit <- glmer(mod, family = "binomial", data = na.omit(dat)) # fit the model
R> tidy(fit, conf.int = TRUE, exponentiate = TRUE)             # print estimates
\end{CodeInput}
\begin{CodeOutput}
# A tibble: 11 x 9
   effect   group term        estim~1 std.er~2 stati~3   p.value conf.~4 conf.~5
   <chr>    <chr> <chr>         <dbl>    <dbl>   <dbl>     <dbl>   <dbl>   <dbl>
 1 fixed    <NA>  (Intercept)  0.0863  0.0182   -11.6   2.99e-31  0.0571   0.130
 2 fixed    <NA>  typeRCT      0.757   0.137     -1.54  1.22e- 1  0.531    1.08 
 3 fixed    <NA>  age          1.03    0.00265   12.9   7.40e-38  1.03     1.04 
 4 fixed    <NA>  motor_scor~  0.651   0.0732    -3.82  1.34e- 4  0.522    0.811
 5 fixed    <NA>  motor_scor~  0.489   0.0555    -6.30  2.97e-10  0.391    0.611
 6 fixed    <NA>  motor_scor~  0.274   0.0321   -11.0   2.28e-28  0.218    0.345
 7 fixed    <NA>  pupilNone    3.20    0.317     11.7   8.18e-32  2.63     3.88 
 8 fixed    <NA>  pupilOne     1.75    0.195      5.06  4.27e- 7  1.41     2.18 
 9 fixed    <NA>  ctIII        2.41    0.268      7.89  3.05e-15  1.94     2.99 
10 fixed    <NA>  ctIV/V       2.30    0.214      8.95  3.55e-19  1.92     2.76 
11 ran_pars name  sd__(Inter~  0.230  NA         NA    NA        NA       NA    
# ... with abbreviated variable names 1: estimate, 2: std.error, 3: statistic,
#   4: conf.low, 5: conf.high
\end{CodeOutput}
\end{CodeChunk}

@@

Create a methods vector and predictor matrix, and make sure
\texttt{name} is not included as predictor, but as clustering variable:

@@

\hypertarget{case-study-iii-hiv-data}{%
\section{Case study III: HIV data}\label{case-study-iii-hiv-data}}

R=region+language+(1\textbar InterviewID) model with with interviewer as
random effects, because the observations are not independent. Interviews
are not allocated randomly. In theory we expect the
inclusion-restriction variable to be randomly assigned, that's why we're
adding region and language to compensate for non-random allocation.

\begin{CodeChunk}
\begin{CodeInput}
R> data("hiv",  package = "GJRM")
R> # We select  5 predictor variables over 9 regions
R> colnames(hiv)
\end{CodeInput}
\begin{CodeOutput}
 [1] "hivconsent"      "hiv"             "age"             "education"      
 [5] "wealth"          "region"          "marital"         "std"            
 [9] "age1sex_cat"     "highhiv"         "partner"         "condom"         
[13] "aidscare"        "knowsdiedofaids" "evertestedHIV"   "smoke"          
[17] "religion"        "ethnicity"       "language"        "interviewerID"  
[21] "sw"             
\end{CodeOutput}
\begin{CodeInput}
R> hivdata <- hiv[,c("hiv","hivconsent","age","marital","condom","highhiv","interviewerID","region","language")]
R> # Study/group variable has to be recoded as integer
R> hivdata$region<-as.integer(hivdata$region)
R> 
R> # Categorical variables have to be recoded as factor
R> # to use the 2l.binary imputation method, it is required that the level names should not have any space between words
R> 
R> hivdata$hiv <- as.factor(hivdata$hiv) #to use the 2l.heckman method, it is required that
R> #the missing variable is stored as a factor in the dataset, otherwise the method will 
R> #apply the imputation correction for a  missing continuous variable instead for 
R> #a missing binary variable,which is in this case the binary response of the hiv test.
R> 
R> hivdata$marital <- as.factor(hivdata$marital)
R> levels(hivdata$marital)<-c("never_married","currently_married","formerly_married")
R> hivdata$condom <- as.factor(hivdata$condom)
R> levels(hivdata$condom)<-c("No_Condom_Last_Intercourse","Condom_Last_Intercourse")
R> hivdata$highhiv <- as.factor(hivdata$highhiv)
R> levels(hivdata$highhiv)<-c("Not_High_Risk_of_HIV","High_Risk_of_HIV")
R> hivdata$interviewerID <- as.factor(hivdata$interviewerID)
R> 
R> hivdata$interviewerID <-  as.factor(as.character(hivdata$interviewerID))
R> interv<-as.data.frame(table(hivdata$interviewerID,hivdata$region))
R> 
R> summary(hivdata)
\end{CodeInput}
\begin{CodeOutput}
   hiv         hivconsent          age                     marital    
 0   :4457   Min.   :0.0000   Min.   :15.00   never_married    :2513  
 1   : 641   1st Qu.:1.0000   1st Qu.:20.00   currently_married:3582  
 NA's:1318   Median :1.0000   Median :28.00   formerly_married : 321  
             Mean   :0.7946   Mean   :30.12                           
             3rd Qu.:1.0000   3rd Qu.:37.00                           
             Max.   :1.0000   Max.   :59.00

                        condom                     highhiv     interviewerID 
 No_Condom_Last_Intercourse:5319   Not_High_Risk_of_HIV:4655   404    : 306  
 Condom_Last_Intercourse   :1097   High_Risk_of_HIV    :1761   303    : 248  
                                                               807    : 247  
                                                               403    : 246  
                                                               903    : 241  
                                                               305    : 232  
                                                               (Other):4896  
     region                   language   
 Min.   :1.000   Language: bemba  :2347  
 1st Qu.:3.000   Language: nyanja :1927  
 Median :5.000   Language: other  : 606  
 Mean   :4.946   Language: lozi   : 604  
 3rd Qu.:7.000   Language: tonga  : 476  
 Max.   :9.000   Language: english: 456  
                 (Other)          :   0  
\end{CodeOutput}
\begin{CodeInput}
R> #  We obtain here the random effects for each interviewer, this is an approximation 
R> #of the interviewer's skill which will be used as an exclusion constraint. 
R> #Here, since the location of the interviewer was not randomly assigned to the 
R> #subjects, the assignment was corrected for region and language. 
R> 
R> hivdata$hivconsent <- as.factor(hivdata$hivconsent)
R> ID_mixed <- lme4::glmer(hivconsent ~ region + language+(1 | interviewerID), data = hivdata, family = binomial)
R> reffect  <- ranef(ID_mixed)$interviewerID
R> reffect$interviewerID <- levels(hivdata$interviewerID)
R> colnames(reffect)     <- c("IDreffect","interviewerID")
R> hivdata  <-  merge(hivdata, reffect, by="interviewerID", all.x=TRUE)
R> 
R> hivdata$interviewerID<-NULL
R> hivdata$hivconsent<-NULL
R> hivdata$language<-NULL
R> summary(hivdata)
\end{CodeInput}
\begin{CodeOutput}
   hiv            age                     marital    
 0   :4457   Min.   :15.00   never_married    :2513  
 1   : 641   1st Qu.:20.00   currently_married:3582  
 NA's:1318   Median :28.00   formerly_married : 321  
             Mean   :30.12                           
             3rd Qu.:37.00                           
             Max.   :59.00                           
                        condom                     highhiv         region     
 No_Condom_Last_Intercourse:5319   Not_High_Risk_of_HIV:4655   Min.   :1.000  
 Condom_Last_Intercourse   :1097   High_Risk_of_HIV    :1761   1st Qu.:3.000  
                                                               Median :5.000  
                                                               Mean   :4.946  
                                                               3rd Qu.:7.000  
                                                               Max.   :9.000  
   IDreffect        
 Min.   :-0.682850  
 1st Qu.:-0.195361  
 Median :-0.080531  
 Mean   :-0.007813  
 3rd Qu.: 0.257165  
 Max.   : 0.667147  
\end{CodeOutput}
\begin{CodeInput}
R> #Set prediction matrix and methods
R> ini <- mice(hivdata, maxit = 0)
R> meth<-ini$method
R> meth["hiv"]<-"2l.binary"
R> pred <- ini$pred
R> pred[,"region"] <- 0
R> pred["region",] <- 0
R> pred["hiv","region"]<- -2
R> pred["hiv","IDreffect"]<- 0
R> 
R> # MAR imputation model, we used the 2l.binary from the miceadds package
R> data_imp_mar <- mice( data= hivdata, # dataset with missing values
+                       m = 10,   # number of imputations
+                       meth = meth, #imputation method vector
+                       pred = pred, #imputation predictors matrix
+                       maxit=1)
\end{CodeInput}
\begin{CodeOutput}

 iter imp variable
  1   1  hiv
  1   2  hiv
  1   3  hiv
  1   4  hiv
  1   5  hiv
  1   6  hiv
  1   7  hiv
  1   8  hiv
  1   9  hiv
  1   10  hiv
\end{CodeOutput}
\begin{CodeInput}
R> # source("mice.impute.2l.heckman.R")
R> # sum(as.numeric(data_imp_mar[["imp"]][["hiv"]][["1"]]))
R> # sum(as.numeric(data_imp_heckp[["imp"]][["hiv"]][["1"]]))
R> # sum(as.numeric(data_imp_mar[["imp"]][["hiv"]][["2"]]))
R> # sum(as.numeric(data_imp_heckp[["imp"]][["hiv"]][["2"]]))
R> 
R> # Heckman model
R> pred["hiv","IDreffect"] <- -3
R> meth<-ini$method
R> meth["hiv"]<-"2l.heckman"
R> 
R> data_imp_heckp <- mice( hivdata, # dataset with missing values
+                         m = 10,   # number of imputations
+                         seed = 1234, #seed attached to the dataID
+                         meth = meth, #imputation method vector
+                         pred = pred, #imputation predictors matrix
+                         print = T,
+                         meta_method="reml",
+                         pmm=FALSE)
\end{CodeInput}
\begin{CodeOutput}

 iter imp variable
  1   1  hiv
  1   2  hiv
  1   3  hiv
  1   4  hiv
  1   5  hiv
  1   6  hiv
  1   7  hiv
  1   8  hiv
  1   9  hiv
  1   10  hiv
  2   1  hiv
  2   2  hiv
  2   3  hiv
  2   4  hiv
  2   5  hiv
  2   6  hiv
  2   7  hiv
  2   8  hiv
  2   9  hiv
  2   10  hiv
  3   1  hiv
  3   2  hiv
  3   3  hiv
  3   4  hiv
  3   5  hiv
  3   6  hiv
  3   7  hiv
  3   8  hiv
  3   9  hiv
  3   10  hiv
  4   1  hiv
  4   2  hiv
  4   3  hiv
  4   4  hiv
  4   5  hiv
  4   6  hiv
  4   7  hiv
  4   8  hiv
  4   9  hiv
  4   10  hiv
  5   1  hiv
  5   2  hiv
  5   3  hiv
  5   4  hiv
  5   5  hiv
  5   6  hiv
  5   7  hiv
  5   8  hiv
  5   9  hiv
  5   10  hiv
\end{CodeOutput}
\end{CodeChunk}
